---
title: "Primers_passos"
author: "Ferran_Pinsach"
date: "February 27, 2019"
output:
  html_document: default
  pdf_document: default
---

# Analysis on Longitudinal Data: Group Project

## Preliminar Study

### Chunk 1: Charging data and summary. RiverSide, numbirths and ID are declared as factors.
```{r}

data<-read.table(paste0(getwd(),"/cattle_mes dades_b.csv"),header=T,fill = TRUE,na.strings="",sep=";")
head(data)
colnames(data)[1]<-"ID"
data<-data[-91,]

summary(data)
colnames(data)

data$Hb<-as.numeric(data$Hb)
data$PCV<-as.numeric(data$PCV)
data$RiverSide<-as.factor(data$RiverSide)
data$numbirths<-as.factor(data$numbirths)
data$ID<-as.factor(data$ID)
data$Time<-as.factor(data$Time)

```
We have observations of ten different cows. For each cow we have 9 differents mesurements, one per timepoint and dose level. 



### Chunk 2: Plots

```{r}
#Grouped data in library nlme
library(nlme)
# 
# data_GD<-groupedData(PCV~Time|ID/Dose,data,outer=~RiverSide,labels=list(y="PCV"))
# 
# #Grouped data-Graphical representation
# 
# plot(data_GD[data_GD$RiverSide==0,],displayLevel = "ID")
# plot(data_GD[data_GD$RiverSide==1,])
# plot(data_GD)
# plot(data_GD,outer=T)
# plot(data_GD[data_GD$Time==1,])
# plot(by(GrowthGV2$height,GrowthGV2$age,mean),type="l",lty=1,lwd=1)
```

```{r}
library(lme4)
library(lattice)

xyplot(PCV~Time,groups=Dose,data,type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))
xyplot(PCV~Time|ID*Dose,data,type="l")

xyplot(PCV~Time,groups=Dose,data[data$Dose=="H",],type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))
xyplot(PCV~Time,groups=Dose,data[data$Dose=="M",],type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))
xyplot(PCV~Time,groups=Dose,data[data$Dose=="L",],type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))

xyplot(PCV~Time,groups=Dose,data[data$RiverSide==1,],type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))
xyplot(PCV~Time,groups=Dose,data[data$RiverSide==0,],type="l",auto.key = list(space = "right", points = FALSE, lines = TRUE))


```

#### First plot: 

1. It seems that PCV value has a positive relation with Dose*Time.
2. Grouping by Dose there is a high variability at intercept level. The intercepts for each dose are mixed.
3. The slopes have high variability too. 
4. There is considerable variability between and within doses. 
5. When RiverSide = 0 there is less variability within groups.

As for each cow we have observations of the PCV for all doses across time so the interactive variable Dose*Time could be considered. On the other hand, in order to fit the variability within groups RiverSide could be considered too.


**Question** For each cow we have 9 observations (3 timpePoints x 3 diff. doses). Are the three mesurements per dose independent between them? Can we consider 9 different timepoints instead of 3x3?

```{r}

data_noNA<-na.omit(data)
(data.lis<-lmList(PCV~Dose|ID/Time,data_noNA))
plot(intervals(data.lis))
```

